---
title: "Assignment 5: Data Visualization"
author: "Nilab Ahmadi"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 First, I'm loading the required packages and load data #how to do cowplot
library(tidyverse) ; library(lubridate) ; library(here); library(cowplot)

# Verifying home directory 
getwd()

#2 Now, I'm reading in the required data 
PeterPaul.chem.nutrients <- 
  read.csv(here 
("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
     stringsAsFactors = TRUE)

Niwot_ridge_litter <- read.csv(here
("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"),
stringsAsFactors = TRUE)

# changing the class of the datasets to date format 

PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
Niwot_ridge_litter$collectDate <- ymd(Niwot_ridge_litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme} 
#3 I'm building a custom theme

# Customize plot background
my_theme = theme_bw() +
theme(
plot.background = element_rect(
fill = 'beige',
color = 'Black'
),
legend.background = element_rect(fill = 'beige'),#legend
axis.line = element_line(
linewidth = 1,
color='Black'
),
axis.text = element_text( #axis labels 
family='serif'
),
plot.title = element_text(  #plot title
face = 'bold',
color='Black',
family='serif'
),#Axis ticks/gridlines
panel.background = element_rect(fill = "beige",colour = "beige",
size = 0.5, linetype = "solid"),)
#Set my theme to the default
theme_set(my_theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 I am loading the dplyr and ggplot2  libraries for data manipulation 
#and data visualization
library(dplyr)
library(ggplot2)

# I'm creating plot for Peter and Paul lakes
ggplot(PeterPaul.chem.nutrients, aes(x=po4, y=tp_ug, color= lakename))+
  geom_point()+
  scale_color_brewer(palette = "Dark2")+
  geom_smooth(method = "lm")+
  xlim(0,50)+
  labs(
x = expression(paste('Phosphate (',mu,'/g',')' )),
y = expression(paste('Total Phosphorus (',mu,'/g)')),
title = 'Q4: Plot of Total Phosphorus vs Phosphate',
color='Lake')
  

```

5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 Here I create  three different box plots for Temperature, 
#TP (Total Phosphorus), and TN (Total Nitrogen)
box_plot <- PeterPaul.chem.nutrients %>%
mutate(the_month = factor(month,levels = 1:12, labels = month.abb)) %>%
ggplot(aes(x=the_month,color=lakename)) #+
scale_x_discrete(labels=month.abb)
 
plt1 <- box_plot+
geom_boxplot(aes(y=temperature_C),show.legend = T) +
labs(x='',y='Temp (C)',title='Temperature') +
theme(legend.position = "top")
 
plt2 <- box_plot+
  geom_boxplot(aes(y=tp_ug),show.legend = F) +
labs(x='',y='Total P',title='Total P')
 
plt3 <- box_plot+
geom_boxplot(aes(y=tn_ug),show.legend = F) +
labs(x='Month',y='Total N',title='Total N',color='Lake')

plot_grid(plt1,plt2,plt3,nrow = 3,align = 'h',rel_heights = c(1.3,1,1))


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: The data shows that both Paul Lake and Peter Lake have similar 
temperature patterns, with temperatures rising in the summer and peaking 
around 20°C from June to August. However, Peter Lake consistently has higher
levels of nutrients, particularly total phosphorus and total nitrogen, compared
to Paul Lake. These differences are most noticeable during the warmer months
(May to September), where Peter Lake's nutrient levels are significantly 
higher, especially in the summer.


6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter 2}
#6 Here I filtered the dataset to include only the "Needles" functional group
#and create the plot

Niwot_ridge_litter %>%
filter(functionalGroup == 'Needles') %>%
ggplot(aes(x=collectDate,y=dryMass,color=nlcdClass)) +
geom_point()


```
```{r Plot litter}

#7 I create the plot with dry mass of needle litter by date, separated 
#by NLCD class using color

Niwot_ridge_litter %>%
filter(functionalGroup == 'Needles') %>%
ggplot(aes(x=collectDate,y=dryMass)) +
geom_point() +
facet_wrap(vars(nlcdClass))


```

Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:Visually,  second plot (No. 7) is usually more effective. 
The color-coded plot (No. 6) shows all the data at once, making it easier 
to see trends quickly. However, if there are many NLCD classes, the lines can
get crowded and hard to read. On the other hand, Plot No. 7, which uses 
separate sections for each class, makes it clearer and easier to understand 
the trends without overlapping lines. This way, each class can be viewed 
individually, helping to highlight differences in dry mass over time. 
While Plot No. 6 is faster for quick comparisons, Plot No. 7 is better 
for a clear and detailed view of the data.