---
title: "Assignment 3: Data Exploration"
author: "Nilab Ahmadi"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r Setup and learning about system}
# I load the necessary packages first
library(tidyverse); library(lubridate); library(here)

#I'm checking the current working directory 
here()

#I'm uploading the two datasets, also renaming them to Neonics and Litter 
Neonics <- read.csv(
  file = here("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv"),
  stringsAsFactors = TRUE
)

Litter <- read.csv(
  file = here("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv"),
  stringsAsFactors = TRUE
    )

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: The neonicotinoids are extensively used class of insecticides 
in agriculture, knowon for their effectiveness against different pests 
ecotoxicology of neonicotinoids is important beacuase of various reasons:
1.Pollinator Health: Neonicotinoids have been linked to adverse effects on 
pollinators, particularly bees. Exposure can impair their foraging behavior
, navigation, and learning abilities, leading to reduced hive productivity 
and increased mortality. Given that pollinators are essential for the 
reproduction of many crops and wild plants, their decline poses a threat to 
biodiversity and agriculture (National Library of medicine). 
Link to the source(https://pubmed.ncbi.nlm.nih.gov/20737791/)
2.These insecticides can leach into water bodies, affecting aquatic insects.
Studies have shown that neonicotinoids are highly toxic to species like 
mayflies and midges, which play vital roles in aquatic ecosystems and serve as 
food sources for other wildlife(National Library of Medicine).
Link to the source 
(https://pmc.ncbi.nlm.nih.gov/articles/PMC4284392/?utm_source=chatgpt.com)

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Stuyding litter and woody debris would be important for undrestanding 
and maintaining the forest ecosystem health. Key reasons include:
1.Habitat Provision: Coarse woody debris (CWD) offers critical habitat for a
variety of organisms, including fungi, invertebrates, amphibians, and small
mammals. For instance, many bird species rely on standing dead trees, or 
snags, for nesting (Wikipedia). Link to the source
(https://en.wikipedia.org/wiki/Snag_%28ecology%29?utm_source=chatgpt.com).
2.Biodiversity Support: The presence of CWD contributes to forest biodiversity
by providing niches for saproxylic (dead wood-dependent) species, which play 
vital roles in decomposition and nutrient cycling(Galen et all., 2018 ).

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

#check the from data-metadata
> Answer: 
 1.Trap-Based Collection System: Litterfall and fine woody debris are 
 collected using elevated and ground traps. Elevated traps
 (0.5m² PVC mesh baskets placed ~80 cm above the ground) 
 collect smaller litterfall (<2 cm in diameter, <50 cm in length), while ground
 traps (3m x 0.5m rectangular areas) capture larger woody debris that is not 
 reliably caught by elevated traps.
 2.Spatial Sampling Design: Sampling occurs at terrestrial NEON sites with 
 woody vegetation >2m tall. Traps are deployed in 20m x 20m or 40m x 40m plots 
 within designated areas, ensuring minimal disturbance to plant diversity and 
 structure measurements. The number of traps varies based on the plot area, 
 with at least one elevated and one ground trap per 400m² plot.
 3. Temporal Sampling Schedule: The frequency of collection varies by 
 ecosystem type. In deciduous forests, traps are sampled biweekly during 
 senescence to capture peak litterfall. In evergreen forests, collection is 
 less frequent (every 1–2 months year-round). Ground traps, which collect 
 fine woody debris, are sampled once per year.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Exploring Data}
# The dimension is [1] 4623   30
dim(Neonics) 

#Getting a quick review of the data
summary(Neonics)

#checking the structure of the data
str(Neonics)

# Viewing the few first rows
head(Neonics)

view(Neonics)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Common Effects of Data}
#running the summary function
Effect_summary <- summary(Neonics$Effect)

#listing the values in a order of magnitude  
sorted_Effects <- sort(Effect_summary, decreasing = TRUE)

# Now, I'm displaying the sorted effects
sorted_Effects

```

> Answer:The effects are tends to be of interest because they provide a 
comprehensive undrestanding of the ecologocial and physiological impact of the 
neonicontinoids on insect population. For example, the population and 
mortality effects are studied to show the decline on insects 
due to neonicontinoids exposure.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Studying the common data}
# I'm determining the species name 
species_summary <- summary(Neonics$Species.Common.Name, maxsum = 6)
#species_summary <- summary(Neonics$Common_Name)

# sorting the species name in descending order
sorted_species <- sort(species_summary, decreasing = TRUE)

# now displaying the top six  most commonly studied species 
sorted_species

```

> Answer:The most commonly studied insect species in the dataset include honey 
bees, parasitic wasps, buff-tailed bumblebees, Carniolan honey bees, 
and bumblebees. These species are vital pollinators and play key roles in 
ecosystem stability and agriculture. Their sensitivity to neonicotinoids makes
them important for research, as their decline threatens biodiversity and 
food production. Studying them helps assess pesticide impacts, guide
conservation efforts, and ensure sustainable pollination and pest control.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r class of the data}
# The class of the "Conc.1..Author." is factor 
class(Neonics$Conc.1..Author.)

```

> Answer:The class of the concentration is factor and not numeric because the 
coulumn contains non-numeric values such as NR, missing data and the format of 
the data is so inconsistent.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Plotting the data}
  
# Now I'm generating the number of studies conducted by publication year
ggplot(Neonics,aes(x=Publication.Year)) +
  geom_freqpoly()

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Creating color graph}
#I'm making the the graph colorful by adding color to the bracket
ggplot(Neonics, aes(x=Publication.Year, color = Test.Location)) +
         geom_freqpoly()

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:The most common testing locations are high-count labs, exceeding
600 counts, and natural field sites, which range from 200 to 400 counts. 
The graph indicates that labs have been the primary testing locations 
since 1990, showing some fluctuations over time, with a significant increase 
noted in 2020, marking its peak. Meanwhile, natural field sites have 
consistently ranked as the second major testing location since 1990.
Between 1990 and 2010, these sites fluctuated at levels higher than 
those of labs, but they subsequently experienced a decline 2010-2020.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Creating bar graph}
# creating a bar graph of endpoints    
ggplot(Neonics) +
geom_bar(aes(x=Endpoint)) +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))   

           
```

> Answer:The two most common endpoints are NOEL and LOEL that exceed the 
1500 count. NOEL is 1816 count and LOEL is 1664 count. Based on appendix, they 
are defined as below:
1.NOEL (No-Observable-Effect-Level): This refers to the highest dose or 
concentration of a substance that does not produce a statistically significant
effect compared to controls.
2.LOEL (Lowest-Observable-Effect-Level): This is the lowest dose or 
concentration of a substance at which there is a significant observed effect 
compared to controls.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Exploring the data}
# I'm checking the class of the collectDate and I see that is a factor not 
#a date
class(Litter$collectDate)

# Now, I'm converting the factor to date 
Litter$collectDate <- as.Date(Litter$collectDate, format='%Y-%m-%d')

# Double checking the class
class(Litter$collectDate)

# Now, I'm checking which dates the Litter was sampled. After running the code,
#I can see the Litter was sampled with dates (2018-08-02 2018-08-30)
unique(Litter$collectDate)

```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Number of Sampled Plots}
# running the code to see how many plot plots were samples at Niwot Ridge
unique(Litter$plotID)

```

> Answer:The difference between unique and summary lies in their function. 
unique retrieves only the distinct values from a dataset, showing which plots 
were sampled without duplication. In contrast, summary provides a statistical 
overview, including the frequency of occurrences for categorical data. 
In this case, the summary output includes counts of different
bee species, indicating how often each category appears rather than 
just listing unique values.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Bar Graph}
#creating the bar graph
ggplot(Litter) +
geom_bar(aes(x=functionalGroup))

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Boxplot and Violin Graphs}
#creating the box plot
ggplot(Litter) +
geom_boxplot(aes(y=dryMass, x=functionalGroup))

#creating the violin plot
ggplot(Litter) +
geom_violin(aes(y=dryMass, x=functionalGroup))


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:The boxplot is more effective because of the data distribution clarity. 
Furthermore,the boxplot is simple and readable. However, the violin plot is not 
as readable as boxplot and the distribution of the data is not also clear.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:Needles and mixed would have the highest biomass at these sites.
